Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2008-10-07 SCP2008/paper/metamonads.tex
changes to code and paper (tex and pdf), as per latest email.

let mo_extend (ip:('a,'b) open_rec) (v:'a) : ('c, 'd) monad = 
  perform s <-- fetch; store (orec_store ip v s)

let mo_lookup (ip:('a,'b) open_rec) : ('c, 'd) monad =
  perform s <-- fetch; ret (orec_find ip s)
\end{code}
%
Currently, we check at generation-time that one should not add an
already existing field to an open record, nor should one attempt to
look up a field that does not exist. It is possible to make these
checks static.  Our approach has the advantage of generating \emph{much}
clearer error messages.

@@

let swap_cols a (n,m) c j =
  let end_vector = n * m in
  let rec loop col_c col_j =
    if col_j < end_vector then
      begin
        swap a col_c col_j;
        loop (col_c + m) (col_j + m)
      end
  in loop c j
\end{code}
We refer to the $a_{ij}$-the element of the matrix as |a.(i*m+j)|,
where |a| is the array component of the record |container2dfromvector|.
The code above defines three auxiliary functions, typically used in 
textbook pseudo-code: |swap a i j| swaps two elements; 
|swap_cols a (n,m) c j| swaps column |c| with column |j| in
the |(n,m)|-matrix |a|. The function 
|swap_rows| |a| |(n,m)| |(r,c) i| swaps row |r| with row |i| in 
the non-yet examined portion of the matrix, 
rectangular block |(r,c)-(n,m)|. We do not touch the 
elements to the left of the column |c| because they are all zeros. 
Since we know the layout of the matrix, we avoid 2D index
computations.

\begin{figure}
\begin{code}[numbers=left,numberblanklines=false,fontsize=\scriptsize]
let find_pivot a (n,m) r c =
  let pivot = ref None in               (* ((i,j), pivot_val) option *)
  begin
  for i = r to n-1 do
    for j = c to m-1 do
      let cur = a.(i*m+j) in
      if not (cur == 0) then
        match !pivot with
        | Some (_,oldpivot) ->
            if abs oldpivot > abs cur then
               pivot := Some ((i,j), cur)
        | None -> pivot := Some ((i,j),cur)
  done; done;
  !pivot
  end
let ge = fun a_orig ->
  let r = ref 0 in                    (* current row index, 0-based *)
  let c = ref 0 in                    (* current column index, 0-based *)
  let a = Array.copy (a_orig.arr) in  (* to avoid clobbering A, save it *)
  let m = a_orig.m in                 (* the number of columns *)
  let n = a_orig.n in                 (* the number of rows *)
  let det_sign = ref 1 in             (* Accumulate sign and magnitude *)
  let det_magn = ref 1 in             (*   of the determinant *)
  while !c < m && !r < n do
    (* Look for a pivot *)
    let pivot = find_pivot a (n,m) !r !c in
    let piv_val = (match  pivot with
    | Some ((piv_r, piv_c),piv_val) ->
        if piv_c <> !c then
          begin
            swap_cols a (n,m) !c piv_c;
            det_sign := - !det_sign (* flip the sign of the det *)
          end;
        if piv_r <> !r then
          begin
            swap_rows a (n,m) (!r,!c) piv_r;
            det_sign := - !det_sign (* flip the sign of the det *)
          end;
        Some piv_val
    | None -> None) in
    (* now do the row-reduction over the (r,c)-(n,m) block *)
    (match piv_val with
    | Some a_rc -> begin
        for ii = !r+1 to n-1 do
          let cur = a.(ii*m + !c) in
          if not (cur == 0)  then
            begin
              for j = !c+1 to m-1 do
                (* fraction-free elimination *)
                a.(ii*m+j) <- (a.(ii*m+j) * a_rc - a.(!r*m+j) * cur) / ! det_magn
              done;
              a.(ii*m+ !c) <- 0
            end;
        done;
        det_magn := a_rc;
        r := !r + 1                 (* advance the rank only if pivot > 0*)
      end
    | None -> det_sign := 0);
    c := !c + 1
    done;
    (* Final result *)
    ({arr=a; n=n; m=m},                 (* The matrix now has the U factor*)
    (if      !det_sign = 0 then  0      (* Compute the signed det *)
     else if !det_sign = 1 then  !det_magn 
     else                        (- !det_magn)),
    !r)                                 (* Rank *)
\end{code}
\caption{Fraction-free in-place Gaussian Elimination over an integer
  matrix}\label{fig:GE-algorithm}
\end{figure}
As is typical of textbook presentations, the main algorithm depends on
a separately-defined function |pivot| to find a pivot. Here we use
full-pivoting, i.e. searching the complete as yet unexamined portion of the
matrix, the rectangular block |(r,c)-(n,m)|, for the element with the 
maximum absolute value. The function returns the value of the
pivot thus found, and its location; the return value is an option type with
|None| indicating that all examined elements are zeroes. In the main
algorithm, after we have found the pivot, we swap the current column 
with the pivot column, and swap the current row with the pivot row (if
necessary). 
After the swap, the $a_{rc}$ element of the matrix is the pivot.
Swapping two rows or two columns changes the sign of the determinant.
After the swaps, the algorithm performs so-called row-reduction
over the |(r,c)-(n,m)| block of the matrix.
We implement the \emph{fraction-free} version of the
algorithm\footnote{sometimes also called the Gauss-Bareiss algorithm}
\cite{Bareiss68},
where the division operation in line 50 assuredly divides two integers
with no remainder, which requires the accumulation
of the determinant.

>>>>

let mo_extend (ip:('a,'b) open_rec) (v:'a) : ('c, unit) monad = 
  perform s <-- fetch; store (orec_store ip v s)

let mo_lookup (ip:('a,'b) open_rec) : ('c, 'a) monad =
  perform s <-- fetch; ret (orec_find ip s)
\end{code}
%
Currently, we check at generation-time that one should not add an
already existing field to an open record, nor should one attempt to
look up a field that does not exist. It is possible to make these
checks static.  Our approach has the advantage of generating \emph{much}
clearer error messages.

@@

let swap_cols a (n,m) c j =
  let end_vector = n * m in
  let rec loop col_c col_j =
    if col_j < end_vector then
      begin
        swap a col_c col_j;
        loop (col_c + m) (col_j + m)
      end
  in loop c j
\end{code}
We refer to the $a_{ij}$\hyp th element of the matrix as |a.(i*m+j)|,
where |a| is the array component of the record |container2dfromvector|.
The code above defines three auxiliary functions, typically used in 
textbook pseudo-code: |swap a i j| swaps two elements; 
|swap_cols a (n,m) c j| swaps column |c| with column |j| in
the |(n,m)|-matrix |a|. The function 
|swap_rows| |a| |(n,m)| |(r,c) i| swaps row |r| with row |i| in 
the non-yet examined portion of the matrix, 
rectangular block |(r,c)-(n,m)|. We do not touch the 
elements to the left of the column |c| because they are all zeros. 
Since we know the layout of the matrix, we avoid 2D index
computations.

\begin{figure}
\begin{code}[numbers=left,numberblanklines=false,fontsize=\scriptsize]
let find_pivot a (n,m) r c =
  let pivot = ref None in               (* ((i,j), pivot_val) option *)
  begin
  for i = r to n-1 do
    for j = c to m-1 do
      let cur = a.(i*m+j) in
      if not (cur == 0) then
        match !pivot with
        | Some (_,oldpivot) ->
            if abs oldpivot > abs cur then
               pivot := Some ((i,j), cur)
        | None -> pivot := Some ((i,j),cur)
  done; done;
  !pivot
  end
let ge = fun a_orig ->
  let r = ref 0 in                    (* current row index, 0-based *)
  let c = ref 0 in                    (* current column index, 0-based *)
  let a = Array.copy (a_orig.arr) in  (* to avoid clobbering A, save it *)
  let m = a_orig.m in                 (* the number of columns *)
  let n = a_orig.n in                 (* the number of rows *)
  let det_sign = ref 1 in             (* Accumulate sign and magnitude *)
  let det_magn = ref 1 in             (*   of the determinant *)
  while !c < m && !r < n do
    (* Look for a pivot *)
    let pivot = find_pivot a (n,m) !r !c in
    let piv_val = (match  pivot with
    | Some ((piv_r, piv_c),piv_val) ->
        if piv_c <> !c then
          begin
            swap_cols a (n,m) !c piv_c;
            det_sign := - !det_sign (* flip the sign of the det *)
          end;
        if piv_r <> !r then
          begin
            swap_rows a (n,m) (!r,!c) piv_r;
            det_sign := - !det_sign (* flip the sign of the det *)
          end;
        Some piv_val
    | None -> None) in
    (* now do the row-reduction over the (r,c)-(n,m) block *)
    (match piv_val with
    | Some a_rc -> begin
        for ii = !r+1 to n-1 do
          let cur = a.(ii*m + !c) in
          if not (cur == 0)  then
            begin
              for j = !c+1 to m-1 do
                (* fraction-free elimination *)
                a.(ii*m+j) <- (a.(ii*m+j) * a_rc - a.(!r*m+j) * cur) / ! det_magn
              done;
              a.(ii*m+ !c) <- 0
            end;
        done;
        det_magn := a_rc;
        r := !r + 1                 (* advance the rank only if pivot > 0*)
      end
    | None -> det_sign := 0);
    c := !c + 1
    done;
    (* Final result *)
    ({arr=a; n=n; m=m},                 (* The matrix now has the U factor*)
    (if      !det_sign = 0 then  0      (* Compute the signed det *)
     else if !det_sign = 1 then  !det_magn 
     else                        (- !det_magn)),
    !r)                                 (* Rank *)
\end{code}
\caption{Fraction-free in-place Gaussian Elimination over an integer
  matrix}\label{fig:GE-algorithm}
\end{figure}
As is typical of textbook presentations, the main algorithm depends on
a separately-defined function |pivot| to find a pivot. Here we use
full-pivoting, i.e. searching the complete as yet unexamined portion of the
matrix, the rectangular block |(r,c)-(n,m)|, for the element with the 
nonzero minimum absolute value. The function returns the value of the
pivot thus found, and its location; the return value is an option type with
|None| indicating that all examined elements are zeroes. In the main
algorithm, after we have found the pivot, we swap the current column 
with the pivot column, and swap the current row with the pivot row (if
necessary). 
After the swap, the $a_{rc}$ element of the matrix is the pivot.
Swapping two rows or two columns changes the sign of the determinant.
After the swaps, the algorithm performs so-called row-reduction
over the |(r,c)-(n,m)| block of the matrix.
We implement the \emph{fraction-free} version of the
algorithm\footnote{sometimes also called the Gauss-Bareiss algorithm}
\cite{Bareiss68},
where the division operation in line 50 assuredly divides two integers
with no remainder, which requires the accumulation
of the determinant.
